---
title: "Mining the Brundy and Thornton (2021) dataset"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---


# Introduction

[Thornton and Brundy (2021)](https://doi.org/10.7710/2162-3309.2410) 
studied the journal prices charged by Elsevier
to five different academic institutions in the USA.  These prices were
negotiated and differed from the [list prices]() available on
Elsevier's web site.  I think this is a valuable study as it shows the
variance paid for titles by different institutions for individual
titles (as opposed to a "big deal" bundle for all of ScienceDirect
content).

The data underlying their paper are available via
[figshare](https://figshare.com/articles/dataset/2020_Elsevier_Journal_Title_Level_Pricing_for_Five_U_S_Research_Universities/13557536).
I have downloaded these files, fixing what I think are a few [minor
errors](data/sjenotes.org).  The data were provided separately for
each institution, so here I reassemble them so that they are arranged
into a table, one row per journal, and columns for each institution.

The key variable is termed *ALP: Adjustment from List Price*.  This is
the percentage difference from list price (negative values meaning a
reduction paid compared to list price).

This interactive document allows you to search by journal, or sort the
data by various columns in the combined dataset.

This interactive document is generated by Rmarkdown, and all source
data are available at <https://github.com/sje30/thornton21>.


```{r init}
require(tidyverse)
require(readr)
require(dplyr)

read_check_data <- function(file, code) {
  ## Read and check the data works out okay.
  debug <- FALSE
  dat <- read_csv(file)
  dat <- rename(dat, published="Published List Price", final="Final Net Price")

  diffs <- dat$final - dat$published
  error1 <- abs(diffs - dat$Difference)

  if (debug && !all( abs(error1)<=0.011))
    browser()

  alp <- (diffs / dat$published) * 100.0
  alp <- round(alp,1)
  error <- abs( alp - dat$ALP)
  if(debug && !all( error < 0.01))
    browser()
  
  dat %>% add_column(institution=code)
  
}
```
 
## Read in the data files

These were cleaned up (notes elsewhere) so that they could be read
into R.  One column seems to be incorrect, which needs fixing.

```{r read-data, echo=TRUE, results='hide',message=FALSE,warning=FALSE}
f <- read_check_data("data/FS_2020_Pricelist2.csv", "FS")
i <- read_check_data("data/ISU_2020_Pricelist2.csv", "ISU")
i$ALP <- i$ALP * -1 #bug -- see below
c <- read_check_data("data/UNC_2020_Pricelist2.csv", "UNC")
a <- read_check_data("data/IA_2020_Pricelist2.csv", "IA")
v <- read_check_data("data/WVU_2020_Pricelist2.csv", "WVU")
dat <- rbind(f, i, c, a, v)
```

Note - bug in the data: ALPs for ISU (Iowa) are the wrong sign.

```

## Plot the ALP for each institution
```{r plot}
ggplot(dat, aes(x=institution, y=ALP, color=institution)) +
  geom_jitter()
```

## Journals with  high ALP

e.g. find those greater than 30% than list price.  No surprises that
the outlier (384% ALP) is the journal *Cell*.

```{r high-alp}
DT::datatable(filter(dat, ALP>=30))
```


##  Journals with low ALP

e.g. find those with at least 30% saving on list price.  
```{r low-alp}
DT::datatable(filter(dat, ALP<=-30))
```


## Raw database of files.

here you can search for individual journals, or sort them by ranking the
different columns.

e.g. by searching for "Trends in Genetics" you can see the variance in
price paid by 3 institutions.

```{r data}
DT::datatable(dat)
```

## Variance of prices

```{r counts}
f1 = f %>% select(Title, published,  FS=ALP)
i1 = i %>% select(Title, published, ISU=ALP)
c1 = c %>% select(Title, published, UNC=ALP)
a1 = a %>% select(Title, published,  IA=ALP)
v1 = v %>% select(Title, published, WVU=ALP)
dat1 <- full_join(f1, i1) %>% full_join(c1) %>% full_join(a1) %>% full_join(v1)
dat2 <- dat1 %>% select(FS, ISU, UNC, IA, WVU) %>% mutate(count=rowSums(!is.na(.)))
dat1$count <- dat2$count
```

Most journals are subscribed to by only one journal:

```{r table-of-counts}
table(dat1$count)
```

This table shows for each journal in the database the price paid by
each institution.  The column 'count' indicates the number of institutions
that subscribed to that journal, and varies between 1 and 5.  

```{r journal-variance}
DT::datatable(dat1)
```

## Things to do

1. Use plotly.R to make figure interactive.
2. Cite reference to the paper.
3. Recompute Table 4 using summary().

